Dynamics of Global and Local Social Influence

### 2021-09-22 manuscript_global_local_influence.Rmd
Structural equivalence & some changes in empirical analysis

---
title: "Dynamics of Global and Local Social Influence"
author: "Fatima Alqabandi & Nicolas Restrepo"
date: '`r format(Sys.time(), "%d %B, %Y")`'
fontsize: 12pt
output:
  pdf_document: default
  html_document: default
header-includes:
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \def\linenumberfont{\normalfont\tiny\sffamily}
bibliography: gli.bib
abstract: "When learning from one another, individuals use different strategies and heuristics. In this paper we argue that the social learning strategies agents use are important for understanding processes of cultural transmission. Even though research on social learning and social networks has increasingly considered different sources of social influence, many models of cultural transmission still tend to assume agents that learn from their peers in a single way. Here, we incorporate current knowledge about social learning into models of cultural transmission and assess its implications. We argue that classic formulations of social influence can be extended to incorporate different social learning heuristics. Then, using longitudinal network data, we show that there is evidence that agents place varying weight on different sources of social information, depending on the phenomenon at hand. We then build an agent-based simulation that incorporates the idea that individuals use different social learning strategies when deciding their lines of action. Our model is capable of replicating interesting findings, including the idea that sub-populations can emerge even in fully connected structures. The key is that we can arrive at that insight only by manipulating the weight that agents place on the different sources of social influence."
---

@@

```{r}
library(tidyverse)
library(igraph)
library(brms)
library(patchwork)
library(panelr)
theme_set(theme_classic())
```

@@

Contemporary conceptions of social influence also wrestle with the tension between global and local dynamics. Increasingly, researchers have moved away from models that only consider immediate ties to more complex conceptualizations that specify varying sources of influence [@dimaggioNetworkEffectsSocial2012]. For the purposes of this analysis, it is important to highlight the work that has shown that central nodes in a structure exercise considerable social influence
76 across the network [@robalinoPeerEffectsAdolescent2018; @paluckSalienceSocialReferents2012]. These findings lend credence to the argument that popular and salient individuals are particularly influential; a view that has been furthered in scholarly work [@sherifReferenceGroupsExploration1964] but that has also been the backbone of policy interventions [@campbellInformalSchoolbasedPeerled2008]. Peer effects then are not monolithic, but rather have varying weights depending on where – and who – they come from.

@@

Multiple strands of research then coalesce around the idea that agents look at different sources when trying the acquire social information. Network researchers have built increasingly complex models of social influence, while scholars interested in social learning provide evidence for the cognitive bases of these learning heuristics. Nonetheless, our models of cultural transmission lag behind. They seldom reflect the fact that agents seek information from different parts of the social structure, that they use different strategies to do so, and that they might give value to such information differently depending on the social phenomena at hand. It remains unclear then how the use of varying strategies of social learning affects processes of cultural transmission.

@@

In order to provide a formal model that specifies the importance of varying sources of influence, we can use as a starting point Friedkin and Johnsen’s [-@friedkinSocialPositionsInfluence1997] classic formulation of social influence. Formalization in this case is useful because it specifies exactly how new ideas of social influence expand upon previous work, and - importantly - the assumptions we have to make in the process. Friedkin and Johnsen’s [-@friedkinSocialPositionsInfluence1997] is as follows:

@@

We focus on four independent variables: drinking, whether a respondent has ran a red light while biking, whether they have copied an illegal CD, and whether they have used public transportation without paying the fare. We choose to examine different variables because our central goal is to analyze whether global and local measures are differentially informative across social phenomena. The questions for these four items were the same: they asked the respondents about their engagement with these behaviors in the last three months. The possible responses were ordinal with five categories: “never,” “one time,” “two to four times,” “five to ten times,” and “more than ten times.” For the purposes of this analysis, we dichotomize the responses so that if a participant answers “never” they get a value of 0 or a value of 1 otherwise. We do this so that the connections between this empirical analysis and the agent-based simulation presented below are clearer.

@@

### Red light ####
bike_model <- read_rds("Analyses/bike_model.rds")

ss <- posterior_samples(bike_model) %>% 
  select(b_friends_z, b_popular_z)

@@

p_bike <- ss_long %>% 
  ggplot(aes(x = coefficient, fill = coef)) +
  geom_density(alpha = 0.5) + 
  labs(x = "", 
       title = "", 
       subtitle = "Red Light") +
  theme(legend.position = "none") +
  theme(legend.title = element_blank())

bike_diff <- ss %>% 
  mutate(difference = b_friends_z-b_popular_z) %>% 
  ggplot(aes(x = difference)) +
  geom_density(fill = "gray80", 
               alpha = 0.5) +
  labs(x = "", 
       y = "", 
       title = "")

@@

### CD ####
copy_model <- read_rds("Analyses/copy_model.rds")

ss <- posterior_samples(copy_model) %>% 
  select(b_friends_z, b_popular_z)

p_copy <- ss_long %>% 
  ggplot(aes(x = coefficient, fill = coef)) +
  geom_density(alpha = 0.5) + 
  labs(x = "", 
       title = "", 
       subtitle = "Copy CD") +
  theme(legend.position = "none") +
  theme(legend.title = element_blank())

copy_diff <- ss %>% 
  mutate(difference = b_friends_z-b_popular_z) %>% 
  ggplot(aes(x = difference)) +
  geom_density(fill = "gray80", 
               alpha = 0.5) +
  labs(x = "", 
       y = "", 
       title = "")

patched <- (p_alc | alc_diff) / (p_bike | bike_diff) /(p_fare | fare_diff) / (p_copy | copy_diff)

@@

Figure 1 shows the marginal posterior distributions for the coefficients for local and global scores for the four models, as well as the difference between the former and the latter. We can read these plots as the predicted increase in the log-odds of incurring in the behavior at hand given a one standard deviation increase in either global or local score. At first glance, we notice that the relative weight of these coefficients is consistently high; an increase in either global or local scores would result in considerably higher log-odds of engaging in the behavior in question. Nonetheless, we also notice subtle differences between local and global dynamics in the models. For illegally copying a CD, the behaviors of the immediate ties and those of the central nodes are almost equally as informative for ego’s behavior in following waves, with the former being slightly more influential. When it comes to running a traffic light and not paying a public transportation fare, however, the distributions of the difference between both coefficients is mostly negative. This implies that increases in the global score would have a bigger effect on the probability of engaging in these behaviors than increases in the local counterpart. We notice the opposite dynamic when we consider drinking: variation in local measures is more informative of one’s drinking habits than the actions of the popular nodes.

We need to be careful when interpreting these coefficients. The coefficients of logit models are notoriously difficult to understand when examined separately. This is because the absolute probability of the outcome depends on all the moving parts of the model, not just the marginal distribution of a single coefficient estimate. Moreover, differences that might seem considerable between high values in the log-odds scale are smaller when transposed to the probability scale. For this reason, it is key to look at the predicted probabilities of engaging in the behaviors in question. Here, we present predicted probabilities for running a red light while biking and for drinking alcohol.

@@

alc <- newdata %>% 
  ggplot(aes(x = type, 
             y = predicted_probs, 
             group = id)) +
  geom_line(alpha = 0.7) + 
  labs(title = "Alcohol", 
       y = "Predicted probability", 
       x = "Type of increase")

@@

dfl_scaled <- read_csv('Data/bike_std.csv')
bike_model <- readRDS('Analyses/bike_model.rds')

@@

bike <- newdata %>% 
  ggplot(aes(x = type, 
             y = predicted_probs, 
             group = id)) +
  geom_line(alpha = 0.7) + 
  labs(title = "Running a red light", 
       y = "Predicted probability", 
       x = "Type of increase", 
       caption = "Figure 2")

(alc | bike)
```

Figure 2 shows two predicted probabilities for each model. We randomly sample 50 agents and predict their probability of engaging in the behavior at hand at different values of the main independent variables. The leftmost point of each plot represents the predicted probability of engaging in the behavior when we add a one standard deviation increase to the global score while keeping local score at its median value. The rightmost point shows the opposite: an increase in local score while holding global score constant. We notice that, in the model that predicts alcohol consumption, there is a clear concentration among lower probabilities; most respondents have a low probability of drinking. However, in many cases, an increase in local score results in a slightly higher probability of engaging in the behavior than a comparable increase in global score. For running a red right while biking, the pattern is reversed. There is a concentration at high probabilities, indicating that most agents have recently ran a red light. The predicted probabilities are slightly higher after a one standard deviation increase in the global score. This means that a rise in global score is more likely to lead to higher predicted probability of engaging in the behavior.

@@

```{r}
set.seed(33)
# Produce initial network visualization
# Create small world network 
network <- sample_smallworld(dim = 1, size = 100, nei = 4, p = 0.1, 10, loops = F)
plot(network, vertex.size = 6, vertex.label = NA, main = "Figure 3")
```

>>>>

---
title: "Dynamics of Global and Local Social Influence"
author: "Nicolas Restrepo & Thomas Wolff"
date: '`r format(Sys.time(), "%d %B, %Y")`'
fontsize: 12pt
output:
  pdf_document: default
  html_document: default
header-includes:
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \def\linenumberfont{\normalfont\tiny\sffamily}
bibliography: gli.bib
abstract: "When learning from one another, individuals use different strategies and heuristics. In this paper we argue that the social learning strategies agents use are important for understanding processes of cultural transmission. Even though research on social learning and social networks has increasingly considered different sources of social influence, many models of cultural transmission still tend to assume agents that learn from their peers in a single way. Here, we incorporate current knowledge about social learning into models of cultural transmission and assess its implications. We argue that classic formulations of social influence can be extended to incorporate different social learning heuristics. Then, using longitudinal network data, we show that there is evidence that agents place varying weight on different sources of social information, depending on the phenomenon at hand. We then build an agent-based simulation that incorporates the idea that individuals use different social learning strategies when deciding their lines of action. Our model is capable of replicating interesting findings, including the idea that sub-populations can emerge even in fully connected structures. The key is that we can arrive at that insight only by manipulating the weight that agents place on the different sources of social influence."
---

@@

```{r}
library(tidyverse)
library(igraph)
library(brms)
library(patchwork)
library(panelr)
library(igraph)
library(ggraph)
library(tidygraph)
theme_set(theme_classic())
```

@@

Contemporary conceptions of social influence also wrestle with the tension between global and local dynamics. Increasingly, researchers have moved away from models that only consider immediate ties to more complex conceptualizations that specify varying sources of influence [@dimaggioNetworkEffectsSocial2012; @burtSocialContagionInnovation1987]. For the purposes of this analysis, it is important to highlight the work that has shown that central nodes in a structure exercise considerable social influence
76 across the network [@robalinoPeerEffectsAdolescent2018; @paluckSalienceSocialReferents2012]. These findings lend credence to the argument that popular and salient individuals are particularly influential; a view that has been furthered in scholarly work [@sherifReferenceGroupsExploration1964] but that has also been the backbone of policy interventions [@campbellInformalSchoolbasedPeerled2008]. Peer effects then are not monolithic, but rather have varying weights depending on where – and who – they come from.

@@

Multiple strands of research then coalesce around the idea that agents look at different sources when trying the acquire social information. Network researchers have built increasingly complex models of social influence, while scholars interested in social learning provide evidence for the cognitive bases of these learning heuristics. What remains unclear, however, is how the use of different learning strategies affects processes of cultural transmission. For instance, recent work has used simulations to explore how beliefs, habits, and the connections between them might emerge and spread across groups. Nonetheless, these models seldom reflect the fact that agents seek information from different parts of the social structure, that they use different strategies to do so, and that they might give value to such information differently depending on the social phenomena at hand.

@@

In order to provide a formal model that specifies the importance of varying sources of influence, we can use as a starting point Friedkin and Johnsen’s [-@friedkinSocialPositionsInfluence1997] classic formulation of social influence. Formalization in this case is useful because it specifies exactly how new ideas of social influence expand upon previous work, and - importantly - the assumptions we have to make in the process. Burt's [-@burtSocialContagionInnovation1987] work remains a paradigmatic example of the usefulness of formalizing different processes of social influence. His formalization of influence through structural equivalence closely  parallels contemporary work on social learning, which shows that agents with domain expertise tend to learn from those who seem similar. In the same vein, it is useful to think how prestige-biased social learning could be incorporated into standard formal models of social influence. Friedkin and Johnsen’s [-@friedkinSocialPositionsInfluence1997] definition is as follows:

@@

We focus on three independent variables: drinking, smoking, and whether the respondents have used public transportation without paying the fare. We choose to examine different variables because our central goal is to analyze whether global and local measures are differentially informative across social phenomena. The questions for these items were the same: they asked the respondents about their engagement with these behaviors in the last three months. The possible responses were ordinal with five categories: “never,” “one time,” “two to four times,” “five to ten times,” and “more than ten times.” For the purposes of this analysis, we dichotomize the responses so that if a participant answers “never” they get a value of 0 or a value of 1 otherwise. We do this so that the connections between this empirical analysis and the agent-based simulation presented below are clearer.

@@

### Smoking ####
smok_model <- read_rds("Analyses/smoke_model.rds")

ss <- posterior_samples(smok_model) %>% 
  select(b_friends_z, b_popular_z)

@@

p_smok <- ss_long %>% 
  ggplot(aes(x = coefficient, fill = coef)) +
  geom_density(alpha = 0.5) + 
  labs(x = "", 
       title = "", 
       subtitle = "Smoking") + 
  theme(legend.position = "top") +
  theme(legend.title = element_blank())

smok_diff <- ss %>% 
  mutate(difference = b_friends_z-b_popular_z) %>% 
  ggplot(aes(x = difference)) +
  geom_density(fill = "gray80", 
               alpha = 0.5) +
  labs(x = "", 
       y = "", 
       title = "")

@@

patched <- (p_alc | alc_diff) / (p_smok | smok_diff) /(p_fare | fare_diff)

@@

Figure 1 shows the marginal posterior distributions for the coefficients for local and global scores for the three models, as well as the difference between the former and the latter. We can read these plots as the predicted increase in the log-odds of incurring in the behavior at hand given a one standard deviation increase in either global or local score. At first glance, we notice that the relative weight of these coefficients is consistently high; an increase in either global or local scores would result in considerably higher log-odds of engaging in the behavior in question. Nonetheless, we also notice subtle differences between local and global dynamics in the models. For smoking, the behaviors of the immediate ties and those of the central nodes are almost equally as informative for ego’s behavior in following waves. When it comes to not paying a public transportation fare, however, the distributions of the difference between both coefficients is mostly negative. This implies that increases in the global score would have a bigger effect on the probability of engaging in these behaviors than increases in the local counterpart. We notice the opposite dynamic when we consider drinking: variation in local measures is more informative of one’s drinking habits than the actions of the popular nodes.

We need to be careful when interpreting these coefficients. The coefficients of logit models are notoriously difficult to understand when examined separately. This is because the absolute probability of the outcome depends on all the moving parts of the model, not just the marginal distribution of a single coefficient estimate. Moreover, differences that might seem considerable between high values in the log-odds scale are smaller when transposed to the probability scale. For this reason, it is key to look at the predicted probabilities of engaging in the behaviors in question. Here, we present predicted probabilities for the three models .

@@

alc <- newdata %>% 
  ggplot(aes(x = type, 
             y = predicted_probs, 
             group = id)) +
  geom_line(alpha = 0.7) + 
  labs(title = "Alcohol", 
       y = "Predicted probability", 
       x = "")

@@

dfl_scaled <- read_csv('Data/fare_std.csv')
bike_model <- readRDS('Analyses/fare_model.rds')

@@

bike <- newdata %>% 
  ggplot(aes(x = type, 
             y = predicted_probs, 
             group = id)) +
  geom_line(alpha = 0.7) + 
  labs(title = "Skipping fare", 
       y = "Predicted probability", 
       x = "", 
       caption = "Figure 2")

#### Alcohol ####
smoke_model <- readRDS('Analyses/smoke_model.rds')
# Import the data 
df_wide <- read_csv('Data/smoke_wide.csv')
# Turn into a long panel dataset 
df_long <- long_panel(df_wide, 
                      id = "id", 
                      prefix = "_w", 
                      begin = 1, 
                      end = 4, 
                      label_location = "end")

# Rename, change, and clean 
dfl <- df_long %>% 
  mutate_at(.vars = c("friends", 
                      "popular", 
                      "actsmok"), 
            as.numeric) %>% 
  filter(!is.na(friends) & 
           !is.nan(friends) & 
           !is.na(popular) & 
           !is.nan(popular))

preds <- predict(smoke_model, 
                 newdata = newdata)

smoke <- newdata %>% 
  ggplot(aes(x = type, 
             y = predicted_probs, 
             group = id)) +
  geom_line(alpha = 0.7) + 
  labs(title = "Smoking", 
       y = "Predicted probability", 
       x = "Type of increase")

(alc | smoke |bike)
```

Figure 2 shows two predicted probabilities for each model. We randomly sample 50 agents and predict their probability of engaging in the behavior at hand at different values of the main independent variables. The leftmost point of each plot represents the predicted probability of engaging in the behavior when we add a one standard deviation increase to the global score while keeping local score at its median value. The rightmost point shows the opposite: an increase in local score while holding global score constant. We notice that, in the model that predicts alcohol consumption, there is a clear concentration among lower probabilities; most respondents have a low probability of drinking. However, in many cases, an increase in local score results in a slightly higher probability of engaging in the behavior than a comparable increase in global score. For using public transportation without paying, the pattern is reversed. There is a concentration at high probabilities, indicating that most agents have recently have skipped such fee. The predicted probabilities are slightly higher after a one standard deviation increase in the global score. This means that a rise in global score is more likely to lead to higher predicted probability of engaging in the behavior.

@@

```{r}
set.seed(33)
# Produce initial network visualization
# Create small world network 
network <- sample_smallworld(dim = 1, size = 100, nei = 4, p = 0.1, 10, loops = F)

graph <- as_tbl_graph(network) %>% 
  mutate(Popularity = centrality_degree(mode = 'in'))

ggraph(graph, 
       layout = "graphopt") + 
  geom_edge_fan(aes(alpha = stat(index)), show.legend = FALSE) + 
  geom_node_point(aes(size = Popularity),color = "steelblue", show.legend = F) + 
  #labs(title = "Figure 3") +
  theme_graph() 
```
